The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-07 manuscript/manuscript.Rnw
add HBE1 figure and discuss opossum

\begin{table}
\tiny
<<top.50.armadillo.expressed.genes,echo=FALSE,results="asis">>=
top.armadillo.genes <- top.genes.table(species.name="dasypus novemcinctus")
print(xtable(top.armadillo.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.armadillo.genes,"Gene symbol")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \armadillo}
\label{tab:top50_armadillo_placenta}
\end{table}

The top 50 expressed genes in the term placenta of \armadillo are in
\cref{tab:top50_armadillo_placenta}.

@@

\begin{table}
  \tiny
<<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
top.opossum.genes <- top.genes.table(species.name="monodelphis domestica")
print(xtable(top.opossum.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.opossum.genes,"Gene symbol")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Monodelphis domestica}}
\label{tab:top50_opossum_placenta}
\end{table}

The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
the opossum are shown in Table 7 with a list of all genes in
Supplemental Table X. The gene with the highest expression
(\ac{FPKM}=41740.6) was hemoglobin, epsilon (HBE), which is expressed at
the embryonic stage of development in marsupials, as in other mammals
(Holland and Gooley, 1997). HBE subunits link up with protein products
of two hemoglobin, zeta (HBZ) genes, which are also expressed to a
high degree in the opossum placenta (\ac{FPKM}=2452.19 and \ac{FPKM}=1139.25).
In addition, there is high expression (\ac{FPKM}=5827.37) of the gene
hemoglobin, omega (HBO), a hemoglobin found only in marsupials and
monotremes (Opazo et al., 2008). The hemoglobin gene HBA is also
expressed in the placenta (\ac{FPKM}=456.978). The gene with the second
highest expression (\ac{FPKM}=22854.8) is superoxide dismutase 3,
extracellular (SOD3), which functions as an antioxidant. Four cystatin
genes (CSTB, CST3, two CSTA-like genes) are expressed to a high
degree, along with two cathepsin-L related genes (Table 7), which are
known to interact with cystatins (Turk et al., 2001). The
uncharacterized gene XM\_001364916.1 also shows high expression in the
opossum placenta (\ac{FPKM}=13572.8). The gene WFDC2, a member of the whey
acidic protein (WAP) family is also expressed to a high degree in the
opossum placenta. Other genes with high expression include genes
involved in retinol binding (RBP4), growth (IGFBP3), lipid metabolism
(APOA1, APOA2), and immunosupression (LGALS1). In addition, there was
little production of CYP19A1 and HSD3B1 (FPKM<2), and moderate
expression of relaxin RLN1 (\ac{FPKM}= 41.3949).

>>>>

\begin{table}
\tiny
<<top.50.armadillo.expressed.genes,echo=FALSE,results="asis">>=
top.armadillo.genes <- top.genes.table(species.name="dasypus novemcinctus")
print(xtable(top.armadillo.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.armadillo.genes,"Gene symbol")
@ 
\caption{\tableexplanation{\armadillo}{tab:all_dasypus_novemcinctus_placenta}}
\label{tab:top50_armadillo_placenta}
\end{table}

The top 50 expressed genes in the term placenta of \armadillo are
shown in \cref{tab:top50_armadillo_placenta} with a list of all genes
with $\mathrm{FPKM}\ge 50$ in Supplemental
\cref{tab:all_dasypus_novemcinctus_placenta}.

@@

\begin{table}
  \tiny
<<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
top.opossum.genes <- top.genes.table(species.name="monodelphis domestica")
print(xtable(top.opossum.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.opossum.genes,"Gene symbol")
@ 
\caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica}}
\label{tab:top50_opossum_placenta}
\end{table}

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \opossum are shown in
\cref{tab:top50_opossum_placenta} with a list of all genes in
Supplemental \cref{tab:all_monodelphis_domestica}. The gene with the
second highest expression is superoxide dismutase 3, extracellular
(\Sexpr{gene.expression("monodelphis_domestica","SOD3",complete=TRUE)}),
which functions as an antioxidant. The gene with the second highest
expression is the hemoglobin subunit epsilon
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000011325",complete=TRUE)}),
which is expressed at the embryonic stage of development in
marsupials, as in other mammals
\cite{Holland.ea1997:Characterizationofembryonicglobin}. \textit{HBE}
subunits link up with protein products of two hemoglobin, zeta
(\textit{HBZ}) genes, which are also expressed to a high degree in the
opossum placenta
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016635",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016631",complete=TRUE)}).
In addition, there is high expression of the gene hemoglobin, omega
(\textit{HBO},
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016637",complete=TRUE)}),
% this is LOC_1000016060 which I figured out by tblastn'ing AAK11486.1
% against Monodelphis domestica
a hemoglobin found only in marsupials and monotremes
\cite{Opazo.ea2008:Genomicevidenceforindependent}. The hemoglobin gene
\textit{HBA} is also expressed in the placenta
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016633",complete=TRUE)}).
Two cystatin genes
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018295",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018296",complete=TRUE)})
are expressed to a high degree, along with one \textit{CTSL} related
gene
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000002844",complete=TRUE)},
\cref{fig:cathespin_evolution_tree}), which are known to interact with
cystatins \cite{Turk.ea2001:Lysosomalcysteineproteasesfacts}. The
uncharacterized gene \textit{LOC10001195}
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000006547",complete=TRUE)})
also shows high expression in the opossum placenta. The gene \textit{WFDC2}, a
member of the whey acidic protein (WAP) family is also expressed to a
high degree in the opossum placenta
(\Sexpr{gene.expression("monodelphis_domestica","WFDC2",complete=TRUE)}).
Other genes with high expression include members of the annexin
complex
(\Sexpr{gene.expression("monodelphis_domestica","S100A6",complete=TRUE)},
\Sexpr{gene.expression("monodelphis_domestica","ANXA2",complete=TRUE)}).
In addition, there was little production of CYP19A1 and HSD3B1
(\Sexpr{gene.expression("monodelphis_domestica","CYP19A1",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000005385",complete=TRUE)},
respectively).

\begin{figure}
<<hbe1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="HBE1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{Expression of \textit{HBE1} and related genes}
\label{fig:hbe1_evolution_tree}
\end{figure}
